---
title: "World Health Organisation: Global Suicide Trends & Analysis"
author: 'Liam Morgan'
date: '01 Apr 2019'
output:
  html_document:
    number_sections: true
    toc: true
    code_folding: "hide"
    theme: readable
    highlight: haddock
---

# **Introduction**

I'm 23 and currently work as an Insight Analyst in the UK.

This is my second kernel - please check out my [most recent kernel here](https://www.kaggle.com/lmorgan95/fraud-detection-smote-xgboost-business-impact), where I use XGBoost to detect credit card fraud (includes business application & cost reduction).

[Feel free to connect on linkedin here :)](https://www.linkedin.com/in/liam-morgan-5168b413a/)



## **Import & Tidy**

The majority of the data used in this analysis was obtained from the World Health Organisation.

***Data Cleaning Notes***

* 7 countries removed (<= 3 years of data total)
* 2016 data was removed (few countries had any, those that did often had data missing)
* HDI was removed due to 2/3 missing data
* Generation variable has problems, detailed in 2.11
* Continent was added to the dataset using the `countrycode` package
* Africa has very few countries providing suicide data

```{r message = FALSE, warning = FALSE}
library(tidyverse) # general
library(ggalt) # dumbbell plots
library(countrycode) # continent
library(rworldmap) # quick country-level heat maps
library(gridExtra) # plots
library(broom) # significant trends within countries

theme_set(theme_light())


# 1) Import & data cleaning

data <- read_csv("../input/master.csv") 

# glimpse(data) # will tidy up these variable names

# sum(is.na(data$`HDI for year`)) # remove, > 2/3 missing, not useable

# table(data$age, data$generation) # don't like this variable

data <- data %>% 
  select(-c(`HDI for year`, `suicides/100k pop`)) %>%
  rename(gdp_for_year = `gdp_for_year ($)`, 
         gdp_per_capita = `gdp_per_capita ($)`, 
         country_year = `country-year`) %>%
  as.data.frame()



# 2) OTHER ISSUES

# a) this SHOULD give 12 rows for every county-year combination (6 age bands * 2 genders):

# data %>% 
#   group_by(country_year) %>%
#   count() %>%
#   filter(n != 12) # note: there appears to be an issue with 2016 data
# not only are there few countries with data, but those that do have data are incomplete

data <- data %>%
  filter(year != 2016) %>% # I therefore exclude 2016 data
  select(-country_year)


# b) excluding countries with <= 3 years of data:

minimum_years <- data %>%
  group_by(country) %>%
  summarize(rows = n(), 
            years = rows / 12) %>%
  arrange(years)

data <- data %>%
  filter(!(country %in% head(minimum_years$country, 7)))


# no other major data issues found yet



# 3) TIDYING DATAFRAME
data$age <- gsub(" years", "", data$age)
data$sex <- ifelse(data$sex == "male", "Male", "Female")


# getting continent data:
data$continent <- countrycode(sourcevar = data[, "country"],
                              origin = "country.name",
                              destination = "continent")

# Nominal factors
data_nominal <- c('country', 'sex', 'continent')
data[data_nominal] <- lapply(data[data_nominal], function(x){factor(x)})


# Making age ordinal
data$age <- factor(data$age, 
                   ordered = T, 
                   levels = c("5-14",
                              "15-24", 
                              "25-34", 
                              "35-54", 
                              "55-74", 
                              "75+"))

# Making generation ordinal
data$generation <- factor(data$generation, 
                   ordered = T, 
                   levels = c("G.I. Generation", 
                              "Silent",
                              "Boomers", 
                              "Generation X", 
                              "Millenials", 
                              "Generation Z"))

data <- as_tibble(data)


# the global rate over the time period will be useful:

global_average <- (sum(as.numeric(data$suicides_no)) / sum(as.numeric(data$population))) * 100000

# view the finalized data
glimpse(data)
```

## **Key Insights**

* Suicide rates are **decreasing globally.** **(2.1)**
* Of those **countries** that show clear linear trends over time, **2/3 are decreasing.** **(2.5.2)**
* On average, **suicide rate increases with age.** **(2.4)**
* This remains true when controlling for continent in the Americas, Asia & Europe, but not for Africa & Oceania. **(2.8)**
* There is a *weak* **positive relationship between a countries GDP (per capita) and suicide rate.** **(2.10)**
* The **highest suicide rate** ever recorded in a demographic (for 1 year) is **225 (per 100k population).** **(2.12)**
* There is an **overrepresentation of men** in suicide deaths at every level of analysis (globally, at a continent and country level). Globally, the male rate is ~3.5x higher. **(2.3) (2.6) (2.7)**


\


# **Global Analysis**

## **Global Trend**

The dashed line is the **global average suicide rate** from 1985 - 2015: **13.1 deaths** (per 100k, per year).

```{r}
data %>%
  group_by(year) %>%
  summarize(population = sum(population), 
            suicides = sum(suicides_no), 
            suicides_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) + 
  geom_line(col = "deepskyblue3", size = 1) + 
  geom_point(col = "deepskyblue3", size = 2) + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  labs(title = "Global Suicides (per 100k)",
       subtitle = "Trend over time, 1985 - 2015.",
       x = "Year", 
       y = "Suicides per 100k") + 
  scale_x_continuous(breaks = seq(1985, 2015, 2)) + 
  scale_y_continuous(breaks = seq(10, 20))
```

***Insights***

* Peak suicide rate was **15.3** deaths per 100k in **1995**
* Decreased steadily, to **11.5** per 100k in **2015** (**~25% decrease**)
* Rates are only now returning to their pre-90's rates
* **Limited data in the 1980's**, so it's hard to say if rate then was truly representative of the global population




## **By Continent**
```{r}
continent <- data %>%
  group_by(continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(suicide_per_100k)

continent$continent <- factor(continent$continent, ordered = T, levels = continent$continent)

continent_plot <- ggplot(continent, aes(x = continent, y = suicide_per_100k, fill = continent)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global Suicides (per 100k), by Continent",
  x = "Continent", 
  y = "Suicides per 100k", 
  fill = "Continent") +
  theme(legend.position = "none", title = element_text(size = 10)) + 
  scale_y_continuous(breaks = seq(0, 20, 1), minor_breaks = F)


continent_time <- data %>%
  group_by(year, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

continent_time$continent <- factor(continent_time$continent, ordered = T, levels = continent$continent)

continent_time_plot <- ggplot(continent_time, aes(x = year, y = suicide_per_100k, col = factor(continent))) + 
  facet_grid(continent ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Continent", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Continent") + 
  theme(legend.position = "none", title = element_text(size = 10)) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)

grid.arrange(continent_plot, continent_time_plot, ncol = 2)
```

***Insights***

* **European rate highest overall**, but has steadily **decreased ~40% since 1995**
* The **European rate for 2015 similar to Asia & Oceania**
* The trendline for Africa is due to poor data quality - just 3 countries have provided data
* **Oceania & Americas trends are more concerning**



## **By Sex**
```{r}
sex_plot <- data %>%
  group_by(sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
ggplot(aes(x = sex, y = suicide_per_100k, fill = sex)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides (per 100k), by Sex",
       x = "Sex", 
       y = "Suicides per 100k") +
  theme(legend.position = "none") + 
  scale_y_continuous(breaks = seq(0, 25), minor_breaks = F)

### with time
sex_time_plot <- data %>%
  group_by(year, sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = factor(sex))) + 
  facet_grid(sex ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Sex", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Sex") + 
  theme(legend.position = "none") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)

grid.arrange(sex_plot, sex_time_plot, ncol = 2)
```

***Insights***

* Globally, the rate of suicide for men has been **~3.5x higher for men**
* Both **male & female suicide rates peaked in 1995**, declining since
* This ratio of 3.5 : 1 (male : female) has remained relatively constant since the mid 90's
* However, during the 80's this ratio was as low as 2.7 : 1 (male : female)

## **By Age**

```{r}
age_plot <- data %>%
  group_by(age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = age, y = suicide_per_100k, fill = age)) + 
  geom_bar(stat = "identity") + 
  labs(title = "Global suicides per 100k, by Age",
       x = "Age", 
       y = "Suicides per 100k") +
  theme(legend.position = "none") + 
  scale_y_continuous(breaks = seq(0, 30, 1), minor_breaks = F)

### with time
age_time_plot <- data %>%
  group_by(year, age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = age)) + 
  facet_grid(age ~ ., scales = "free_y") + 
  geom_line() + 
  geom_point() + 
  labs(title = "Trends Over Time, by Age", 
       x = "Year", 
       y = "Suicides per 100k", 
       color = "Age") + 
  theme(legend.position = "none") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F)


grid.arrange(age_plot, age_time_plot, ncol = 2)
```

***Insights***

* Globally, the **likelihood of suicide increases with age**
* Since 1995, suicide rate for everyone aged **>= 15** has been **linearly decreasing**
* The suicide rate of those aged **75+** has dropped by more than **50%** since 1990
* Suicide rate in the **'5-14'** category remains **roughly static** and small (< 1 per 100k per year)


## **By Country**

### **Overall**
```{r fig.width = 7, fig.height = 12}
country <- data %>%
  group_by(country, continent) %>%
  summarize(n = n(), 
            suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  arrange(desc(suicide_per_100k))

country$country <- factor(country$country, 
                          ordered = T, 
                          levels = rev(country$country))

ggplot(country, aes(x = country, y = suicide_per_100k, fill = continent)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  labs(title = "Global suicides per 100k, by Country",
       x = "Country", 
       y = "Suicides per 100k", 
       fill = "Continent") +
  coord_flip() +
  scale_y_continuous(breaks = seq(0, 45, 2)) + 
  theme(legend.position = "bottom")
```

***Insights***

* Lithuania's rate has been highest by a large margin: > 41 suicides per 100k (per year)
* Large overrepresentation of European countries with high rates, few with low rates


\


Below is a geographical heat map of the suicide rates between the timeframe of this analysis - **note the lack of data for Africa and Asia**, and bear in mind that 7 countries have been removed due to insufficient data.

```{r message = FALSE, warning = FALSE, results = 'hide', fig.width = 7}
country <- data %>%
  group_by(country) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

countrydata <- joinCountryData2Map(country, joinCode = "NAME", nameJoinColumn = "country")

par(mar=c(0, 0, 0, 0)) # margins

mapCountryData(countrydata, 
nameColumnToPlot="suicide_per_100k", 
mapTitle="", 
colourPalette = "heat", 
oceanCol="lightblue", 
missingCountryCol="grey65", 
catMethod = "pretty")

mapCountryData(countrydata, 
nameColumnToPlot="suicide_per_100k", 
mapTitle="", 
mapRegion = "eurasia", 
colourPalette = "heat", 
oceanCol="lightblue", 
missingCountryCol="grey65", 
addLegend = FALSE, 
catMethod = "pretty")
```

It's important to note that looking at figures at a global/continent level might not truly be representative of the globe/continent for these reasons.

Comparing the raw suicide rates of countries may also lead to some issues - the definition of suicide (and the reliability that a death is recorded as suicide) will likely vary between countries. 

However, trends over time (within countries) are likely to be reliable. I address this next.




### **Linear Trends**

I'm interested in how the suicide rate is changing over time within each country. Instead of visualizing all 93 countries rates across time, I fit a simple linear regression to every countries data. I extract those with a 'year' p-value (*corrected for multiple comparisons*) of < 0.05. 

In other words: as time goes on, **I look for countries where the suicide rate is linearly increasing or decreasing over time**. These can then be rank ordered by their 'year' coefficient, which would be their rate of change as time goes on.

```{r}
country_year <- data %>%
  group_by(country, year) %>%
  summarize(suicides = sum(suicides_no), 
            population = sum(population), 
            suicide_per_100k = (suicides / population) * 100000, 
            gdp_per_capita = mean(gdp_per_capita))


country_year_trends <- country_year %>%
  ungroup() %>%
  nest(-country) %>% # format: country, rest of data (in list column)
  mutate(model = map(data, ~ lm(suicide_per_100k ~ year, data = .)), # for each item in 'data', fit a linear model
         tidied = map(model, tidy)) %>% # tidy each of these into dataframe format - call this list 'tidied'
  unnest(tidied)

country_year_sig_trends <- country_year_trends %>%
  filter(term == "year") %>%
  mutate(p.adjusted = p.adjust(p.value, method = "holm")) %>%
  filter(p.adjusted < .05) %>%
  arrange(estimate)

country_year_sig_trends$country <- factor(country_year_sig_trends$country, 
                                          ordered = T, 
                                          levels = country_year_sig_trends$country)
```


```{r fig.width = 7, fig.height = 8}
# plot 1
ggplot(country_year_sig_trends, aes(x=country, y=estimate, col = estimate)) + 
  geom_point(stat='identity', size = 4) +
  geom_hline(yintercept = 0, col = "grey", size = 1) +
  scale_color_gradient(low = "green", high = "red") +
  geom_segment(aes(y = 0, 
                   x = country, 
                   yend = estimate, 
                   xend = country), size = 1) +
  labs(title="Change per year (Suicides per 100k)", 
       subtitle="Of countries with significant trends (p < 0.05)", 
       x = "Country", y = "Change Per Year (Suicides per 100k)") +
  scale_y_continuous(breaks = seq(-2, 2, 0.2), limits = c(-1.5, 1.5)) +
  theme(legend.position = "none") +
  coord_flip()
```

***Insights***

* **~1/2** of all countries suicide rates are **changing linearly** as time progresses
* 32 (**2/3**) of these 48 countries are **decreasing**
* Overall, this is painting a positive picture


\


Steepest *increasing* trends:

```{r}
### Lets look at those countries with the steepest increasing trends

top12_increasing <- tail(country_year_sig_trends$country, 12)

country_year %>%
  filter(country %in% top12_increasing) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ country) + 
  theme(legend.position = "none") + 
  labs(title="12 Steepest Increasing Trends", 
       subtitle="Of countries with significant trends (p < 0.05)", 
       x = "Year", 
       y = "Suicides per 100k")
```

***Insights***

* **South Korea** shows the most concerning trend - an increase in suicide of 0.931 people (per 100k, per year) - the **steepest increase globally**
* **Guyana** is similar, at + 0.925 people (per 100k, per year)
* Between **1998 and 1999 (5.3 to 24.8)**, Guyana's rate increased by ~365%
* The historical data for Guyana seems questionable - it's known for very high suicide rates but the jump seems unlikely (maybe changed how they classified suicide?)


\


Steepest *decreasing* trends:


```{r}
### Now those with the steepest decreasing trend

top12_decreasing <- head(country_year_sig_trends$country, 12)

country_year %>%
  filter(country %in% top12_decreasing) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ country) + 
  theme(legend.position = "none") + 
  labs(title="12 Steepest Decreasing Trends", 
       subtitle="Of countries with significant trends (p < 0.05)", 
       x = "Year", 
       y = "Suicides per 100k")
```

***Insights***

* **Estonia** shows the most positive trend - **every year, ~1.31 *less* people (per 100k) commit suicide** - the steepest decrease globally
* Between 1995 and 2015, this drops from **43.8 to 15.7** per 100k (per year) - a **64% decrease**
* The Russian Federation trend is interesting, only beginning to drop in 2002. Since then it has decreased by ~50%.

## **Gender differences, by Continent**


```{r}
data %>%
  group_by(continent, sex) %>%
  summarize(n = n(), 
            suicides = sum(as.numeric(suicides_no)), 
            population = sum(as.numeric(population)), 
            suicide_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = continent, y = suicide_per_100k, fill = sex)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  labs(title = "Gender Disparity, by Continent",
   x = "Continent", 
   y = "Suicides per 100k", 
   fill = "Sex") +
  coord_flip()
```

***Insights***

* **European men** were at the **highest risk** between 1985 - 2015, at ~ 30 suicides (per 100k, per year)
* **Asia** had the smallest overrepresentation of male suicide - the rate was **~2.5x as high for men**
* Comparatively, **Europe's** rate was **~3.9x as high for men**


## **Gender differences, by Country**

```{r fig.width = 7, fig.height = 12}
country_long <- data %>%
  group_by(country, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  mutate(sex = "OVERALL")

### by country, continent, sex

sex_country_long <- data %>%
  group_by(country, continent, sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)


sex_country_wide <- sex_country_long %>%
  spread(sex, suicide_per_100k) %>%
  arrange(Male - Female)


sex_country_wide$country <- factor(sex_country_wide$country, 
                                         ordered = T, 
                                         levels = sex_country_wide$country)

sex_country_long$country <- factor(sex_country_long$country, 
                                   ordered = T, 
                                   levels = sex_country_wide$country) # using the same order



### this graph shows us how the disparity between deaths varies across gender for every country
# it also has the overall blended death rate - generally countries with a higher death rate have a higher disparity
# this is because, if suicide is more likely in a country, the disparity between men and women is amplified

ggplot(sex_country_wide, aes(y = country, color = sex)) + 
  geom_dumbbell(aes(x=Female, xend=Male), color = "grey", size = 1) + 
  geom_point(data = sex_country_long, aes(x = suicide_per_100k), size = 3) +
  geom_point(data = country_long, aes(x = suicide_per_100k)) + 
  geom_vline(xintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  theme(axis.text.y = element_text(size = 8), 
        legend.position = c(0.85, 0.2)) + 
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  labs(title = "Gender Disparity, by Continent & Country", 
       subtitle = "Ordered by difference in deaths per 100k.", 
       x = "Suicides per 100k", 
       y = "Country", 
       color = "Sex")
```


```{r fig.width = 7, fig.height = 12}
country_gender_prop <- sex_country_wide %>%
  mutate(Male_Proportion = Male / (Female + Male)) %>%
  arrange(Male_Proportion)

sex_country_long$country <- factor(sex_country_long$country, 
                                   ordered = T,
                                   levels = country_gender_prop$country)

ggplot(sex_country_long, aes(y = suicide_per_100k, x = country, fill = sex)) + 
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportions of suicides that are Male & Female, by Country", 
       x = "Country", 
       y = "Suicides per 100k",
       fill = "Sex") + 
  coord_flip()
```

***Insights***

* The **overrepresentation of men** in suicide deaths appears to be **universal**, and can be observed to differing extents in every country
* Whilst **women are more likely to suffer from depression and suicidal thoughts, men are more likely to die from suicide**
* This is known as the [gender paradox on suicidal behaviour](https://en.wikipedia.org/wiki/Gender_differences_in_suicide)

## **Age differences, by Continent**

```{r}
data %>%
  group_by(continent, age) %>%
  summarize(n = n(), 
            suicides = sum(as.numeric(suicides_no)), 
            population = sum(as.numeric(population)), 
            suicide_per_100k = (suicides / population) * 100000) %>%
  ggplot(aes(x = continent, y = suicide_per_100k, fill = age)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_hline(yintercept = global_average, linetype = 2, color = "grey35", size = 1) +
  labs(title = "Age Disparity, by Continent",
       x = "Continent", 
       y = "Suicides per 100k", 
       fill = "Age")
```

***Insights***

* For the **Americas, Asia & Europe** (which make up most of the dataset), **suicide rate increases with age**
* **Oceania & Africa's rates are highest for those aged 25 - 34**



## **As a country gets richer, does it's suicide rate decrease?**

**It depends on the country** - for almost every country, there is a high correlation between `year` and `gdp per capita`, i.e. as time goes on, gdp per capita linearly increases.

```{r}
country_year_gdp <- data %>%
  group_by(country, year) %>%
  summarize(gdp_per_capita = mean(gdp_per_capita))
  
country_year_gdp_corr <- country_year_gdp %>%
  ungroup() %>%
  group_by(country) %>%
  summarize(year_gdp_correlation = cor(year, gdp_per_capita))
```

I calculated the pearson correlations between 'year' and 'GDP per capita' within each country, then summarized the results:

The mean correlation was **`r round(mean(country_year_gdp_corr$year_gdp_correlation), 3)`**, indicating a *very* strong positive linear relationship. 


This basically means that looking within a country and asking **"does an increase in weath (per person) have an effect suicide rate"** is pretty similar to asking **"does a countries suicide rate increase as time progresses"**.

This was answered earlier in **(2.5.2)** - it depends on the country! Some countries are increasing with time, most are decreasing.


Instead, I ask a slightly different question below.



## **Do richer countries have a higher rate of suicide?**

Instead of looking at trends within countries, here I take every country and calculate their mean GDP (per capita) across all the years in which data is available. I then measure how this relates to the countries suicide rate across all those years.

The end result is one data point per country, intended to give a general idea of the wealth of a country and its suicide rate.

```{r}
country_mean_gdp <- data %>%
  group_by(country, continent) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000, 
            gdp_per_capita = mean(gdp_per_capita))

ggplot(country_mean_gdp, aes(x = gdp_per_capita, y = suicide_per_100k, col = continent)) + 
  geom_point() + 
  scale_x_continuous(labels=scales::dollar_format(prefix="$"), breaks = seq(0, 70000, 10000)) + 
  labs(title = "Correlation between GDP (per capita) and Suicides per 100k", 
       subtitle = "Plot containing every country",
       x = "GDP (per capita)", 
       y = "Suicides per 100k", 
       col = "Continent") 
```

There are quite a few high leverage & residual countries that could have a significant impact on the fit of my regression line (e.g. Lithuania, top left). I'll identify and exclude these using Cooks Distance, excluding those countries with a CooksD value of greater than 4/n.

I assess the statistics of this model (with outliers removed) below.

```{r}
model1 <- lm(suicide_per_100k ~ gdp_per_capita, data = country_mean_gdp)

gdp_suicide_no_outliers <- model1 %>%
  augment() %>%
  arrange(desc(.cooksd)) %>%
  filter(.cooksd < 4/nrow(.)) %>% # removes 5/93 countries
  inner_join(country_mean_gdp, by = c("suicide_per_100k", "gdp_per_capita")) %>%
  select(country, continent, gdp_per_capita, suicide_per_100k)

model2 <- lm(suicide_per_100k ~ gdp_per_capita, data = gdp_suicide_no_outliers)

summary(model2)
```

The **p-value** of the model is **`r round(glance(model2)$p.value, 4)`** < 0.05. This means we can *reject* the hypothesis that a countries GDP (per capita) has no association with it's rate of suicide (per 100k).

The r-squared is **`r round(glance(model2)$r.squared, 4)`**, so GDP (per capita) explains very little of the variance in suicide rate overall.

**What does all this mean?**

There is a weak but significant positive linear relationship - ***richer* countries are associated with *higher* rates of suicide**, but this is a ***weak* relationship** which can be seen from the graph below.

```{r}
ggplot(gdp_suicide_no_outliers, aes(x = gdp_per_capita, y = suicide_per_100k, col = continent)) + 
  geom_point() + 
  geom_smooth(method = "lm", aes(group = 1)) + 
  scale_x_continuous(labels=scales::dollar_format(prefix="$"), breaks = seq(0, 70000, 10000)) + 
  labs(title = "Correlation between GDP (per capita) and Suicides per 100k", 
       subtitle = "Plot with high CooksD countries removed (5/93 total)",
       x = "GDP (per capita)", 
       y = "Suicides per 100k", 
       col = "Continent") + 
  theme(legend.position = "none")
```

This line of best fit is represented by the equation below, where:

* Suicides = Suicides per 100k
* GDP = GDP per capita (in thousands, USD)

$$ Suicides = 8.7718 + 0.1115*GDP $$

This means that, at a country level and over the time frame of this analysis (1985 - 2015), **an increase of GDP (per capita) by $8,967 was associated with 1 additional suicide, per 100k people, per year**.



## **The reason I haven't used the generation variable**

With *continuous data*, if you have someones age in a given year, you have their generation. The graph below demonstrates how this works for this dataset really well, and is equivalent to the graph of age across time, shown in **(2.4)**.

```{r fig.height = 8}
data %>%
  group_by(generation, age, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = factor(generation, ordered = F))) + 
  geom_point() + 
  geom_line() + 
  facet_grid(age ~ ., scales = "free_y") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = NULL) +
  labs(title = "Relationship between Generation, Age & Year", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Generation") + 
  theme(legend.position = "bottom")
```

However, because of the overlap of different age categories, trying to interpret the trend of generation suicide rates over time creates problems. 

Compare the rates below to the plot above - **large spikes** occur at the same time that different age bands begin/stop being classified as from a certain generation.

Note, for example, the **supposed spike in suicide rate for G.I. generation in 1991**, where those aged '55 - 75' suddenly stop being classified as from this generation.

```{r fig.height = 9}
generation_rate <- data %>%
  group_by(generation, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = factor(generation, ordered = F))) + 
  geom_point() + 
  geom_line() + 
  facet_grid(generation ~ ., scales = "free_y") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = NULL) +
  labs(title = "Suicides per 100k, by Generation", 
       x = "Year", 
       y = "Suicides per 100k") + 
  theme(legend.position = "none")

generation_population <- data %>%
  group_by(generation, year) %>%
  summarize(population = sum(as.numeric(population))) %>%
  ggplot(aes(x = year, y = population / 1000000, col = factor(generation, ordered = F))) + 
  geom_point() + 
  geom_line() + 
  facet_grid(generation ~ ., scales = "free_y") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = NULL) +
  labs(title = "Population, by Generation", 
       x = "Year", 
       y = "Population (Millions)", 
       col = "Generation") + 
  theme(legend.position = "none")

grid.arrange(generation_rate, generation_population, ncol = 2)

```

This is probably a **problem with how the dataset was created** - it looks like the generation variable was created after the data was summarized (by country, year, age, sex) and just appended onto the end. This *shouldn't be possible*, because not everyone in a given age band & year will be of one generation.

This shows why the 'spikes' in generation across time are pretty meaningless and I would **recommend to others not to use the variable**, as it can will lead to wrong conclusions.


## **The 5% highest risk instances in history**

I'll filter out data from 1985 only and look at what happens in the 3 decades following.

Here i'm interested in the 5% highest risk (suicides/100k) demographics between 1986 and 2015.

I define a *demographic* as a **year** in a particular **country**, for some combination of **sex** & **age**. e.g. 'United Kingdom, 2010, Female, 15 - 24' would be a single demographic/point on the jitter plot below.

In order for a demographic to be in the top 5% for historic suicide rates, it would require a suicide rate exceeding 50.7 (per 100k) in that year.

```{r}
demographic_most <- data %>%
  mutate(suicides_per_100k = suicides_no * 100000 / population) %>%
  arrange(desc(suicides_per_100k)) %>% 
  filter(year != 1985) %>%
  head(n = round(nrow(.) * 5 / 100))
  

demographic_most$time <- ifelse(demographic_most$year <= 1995, "1986 - 1995", 
                                ifelse(demographic_most$year <= 2005, "1996 - 2005", 
                                       "2006 - 2015"))
```

```{r}
ggplot(demographic_most, aes(x = age, fill = sex)) + 
  geom_bar() + 
  labs(title = "5% Most At-Risk Instances in History", 
       subtitle = "Volumes by Decade, Age & Sex",
       x = "Age", 
       y = "Number of Demographics", 
       fill = "Sex") + 
  facet_wrap(~ time) + 
  scale_y_continuous(breaks = seq(0, 300, 20))

set.seed(1)

ggplot(demographic_most, aes(x = age, y = suicides_per_100k, col = sex)) + 
  geom_jitter(alpha = 0.5) + 
  labs(title = "5% Most At-Risk Instances in History", 
       subtitle = "Instances by Decade, Age, & Sex",
       x = "Age", 
       y = "Suicides per 100k", 
       col = "Sex") + 
  facet_wrap(~ time) + 
  scale_y_continuous(breaks = seq(50, 300, 10))
```

***Insights***

* **44.5%** of these 'high risk' instances occurred **between 1996 and 2005**
* **53.5%** were in the **75+** age category
* **96.9% were a male demographic**
* Of the 3.1% (42 instances) that *were* for **women**, **41/42** of these were in the **75+** demographic
* The **highest suicide rate** for a demographic in any year is **225 (per 100k)** -  that's 0.225% of the entire demographic committing suicide in 1 year



Two of the most **consistently at-risk** demographics seem to be **men** in **South Korea & Hungary**, which I will visualize below.

```{r}
data %>%
  filter(country %in% c('Republic of Korea', 'Hungary'), sex == "Male") %>%
  group_by(country, age, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_line() + 
  geom_point() + 
  facet_wrap(~ age) + 
  geom_hline(yintercept = min(demographic_most$suicides_per_100k)) + 
  theme(legend.position = "bottom") + 
  scale_y_continuous(breaks = seq(0, 220, 40)) +
  labs(title = "Male Age-Group Trends in Hungary & South Korea", 
       subtitle = "Black reference line indicates where the demographic enters the 'top 5% in history'",
       x = "Year", 
       y = "Suicides per 100k",
       col = "Country")
```

Two very different trends emerge. Hungary is obviously moving in a positive direction, whereas South Korea appears to be coming out of somewhat of a crisis. 

For **South Korea**, **mens rates in the 75+ category** increased from 26.2 (per 100k) in 1992, to a peak of 185 (per 100k) in 2011 - **an *increase* of more than 600%**. Men aged 55-74 see a similar increase.

This was highlighted by my statistical analysis in **(2.5.2)**, which identified South Korea as the steepest increasing country, and Hungary as the 4th steepest decreasing country overall.



\


# **Comparing the UK, Ireland, America, France & Denmark**

I think it would be useful to compare a few countries that people might think of as similar to the UK (culturally, legally, economically).


## **Overall Trend**

```{r message = FALSE, warning = FALSE}
data_filtered <- data %>%
  filter(country %in% c("United Kingdom", 
                        "Ireland",
                        "United States", 
                        "France", 
                        "Denmark")) 


data_filtered %>%
  group_by(country, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = F, span = 0.2) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + 
  labs(title = "UK, Ireland, US, France & Denmark", 
       subtitle = "Suicides per 100k population, 1985 - 2015", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Country")
```

***Insights***

* The **UK suicide rate has been consistently lowest since 1990**, and has remained fairly static since ~1995
* France has historically had the highest rate, but is now roughly equal with America
* The **US has the most concerning trend, linearly increasing by ~1/3 since 2000**


## **By Sex**

### **Male & Female Rates (over time)**

```{r fig.height = 8, message = FALSE, warning = FALSE}
data_filtered %>%
  group_by(country, sex, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = F, span = 0.2) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + 
  facet_wrap(~ sex, scales = "free_y", nrow = 2) +
  labs(title = "UK, Ireland, US, France & Denmark", 
       subtitle = "Suicides per 100k population, 1985 - 2015", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Country")
```

***Insights***

* For the **UK**, there's **no obvious increase in the suicide rate for men than can't also be observed to an equal extent in women**
* Again, for men and women, **France has decreased** to being roughly equal with the US in 2015
* The different trend lines for men & women in **Ireland** is unusual - in **1990**, the **male rate increases**, but the **same can't be observed for females**


## **2010 - 2015 Only**

For the purposes of these visualisations, i'm really more interested in data from recent years (France, for example, has changed a *lot*), so i'll **restrict the timeframe to 2010 onwards**.

### **Proportion of suicides that are Men**

```{r}
t1 <- data_filtered %>%
  filter(year >= 2010) %>%
  group_by(sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

global_male_proportion <- t1$suicide_per_100k[2] / sum(t1$suicide_per_100k)


t2 <- data_filtered %>%
  filter(year >= 2010, continent == "Europe") %>%
  group_by(sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000)

european_male_proportion <- t2$suicide_per_100k[2] / sum(t2$suicide_per_100k)


data_filtered %>%
  filter(year >= 2010) %>%
  group_by(country, sex) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = country, y = suicide_per_100k, fill = sex)) + 
  geom_bar(position = "fill", stat = "identity") + 
  geom_hline(yintercept = global_male_proportion) + 
  geom_hline(yintercept = european_male_proportion, col = "blue") + 
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "Proportion of suicides that were Male & Female", 
       subtitle = "2010 - 2015 only, with reference lines for Europe (blue) & Globally (black)", 
       x = "Country", 
       y = "", 
       fill = "Sex")
```

***Insights***

* Similar pattern as seen throughout the analysis - **men make up ~ 75% of deaths by suicide**
* The **highest** proportion is in **Ireland - 81.7% male**
* The **lowest** proportion is for **Denmark - 73.5% male**


### **Age Rates**

```{r}
data_filtered %>%
  filter(year >= 2010) %>%
  group_by(country, age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = country, y = suicide_per_100k, fill = age)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Age ", 
       subtitle = "2010 - 2015 only", 
       x = "Country", 
       y = "Suicides per 100k", 
       fill = "Age")
```


***Insights***

* There's a huge difference in the 'trend' of suicide rates as age varies within each country
* **Suicide rate increases with age for France, Denmark and the US** (to a lesser extent)
* Those aged **35-54** at the **highest risk** in **Ireland and the UK**, which follow closer to a gaussian distribution

### **Male & Female Rates (for different age categories)**

```{r fig.width = 7, fig.height = 8}

data_filtered %>% 
  filter(year >= 2010) %>%
  group_by(country, sex, age) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = age, y = suicide_per_100k, fill = country)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  facet_wrap(~ sex, scales = "free_x") +
  labs(title = "Age Disparity, by Country",
       subtitle = "2010 - 2015 only",
       x = "Age", 
       y = "Suicides per 100k", 
       fill = "Country") +
  coord_flip() + 
  theme(legend.position = "bottom")
```

***Insights***

* In the **US**, **suicide rate for men continues to increase with age**, but the **female rate decreases in old age**
* This weird disparity is only present in the US and i'm curious as to why it occurs
* The **UK has the lowest or second lowest suicide rate in every sex-age group**



## **Young to Middle-Aged Men**

There is a big concern in my country (UK) regarding mental health problems and suicide for young to middle-aged men. Here i'm going to **restrict the analysis** to just:

* **Men**
* **Ages "15-24", "25-34" & "35-54"**

I'll basically be observing whether concerning trends are present. I think having other countries here for comparison will be useful and will help provide perspective in the analysis.

### **Men - Ages 15-54 Combined**

```{r message = FALSE, warning = FALSE}
data_filtered %>%
  filter(age %in% c("15-24", "25-34", "35-54"), sex == "Male") %>%
  group_by(country, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = F, span = 0.2) + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + 
  labs(title = "UK, Ireland, US, France & Denmark", 
       subtitle = "Suicides per 100k population, 1985 - 2015", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Country")
```

***Insights***

* **Ireland's trend over the 1990's was very concerning**
* It went from 14 (per 100k, per year) to 33.3 between 1988 and 1998 - an **increase of 138%**
* Again, the **US** shows the most obvious and **concerning current trend**
* *Comparatively*, for young to middle-aged men, the **UK seems fairly flat across time**


### **Men - Ages 15-24, 25-34 & 35-54**

```{r fig.height = 12, message = FALSE, warning = FALSE}
data_filtered %>%
  filter(age %in% c("15-24", "25-34", "35-54"), sex == "Male") %>%
  group_by(country, age, year) %>%
  summarize(suicide_per_100k = (sum(as.numeric(suicides_no)) / sum(as.numeric(population))) * 100000) %>%
  ggplot(aes(x = year, y = suicide_per_100k, col = country)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(se = F, span = 0.2) + 
  facet_wrap(~ age, nrow = 3, scales = "free_y") + 
  scale_x_continuous(breaks = seq(1985, 2015, 5), minor_breaks = F) + 
  labs(title = "UK, Ireland, US, France & Denmark", 
       subtitle = "Suicides per 100k population, 1985 - 2015", 
       x = "Year", 
       y = "Suicides per 100k", 
       col = "Country")
```

***Insights***

* For **men in the UK**, only the **'35-54'** category seems to be increasing, with a **slight increase (~10-15%)** over the past decade
* **UK** rates for men in the **'15-24'** and **'25-34'** categories appear **flat** & **slightly decreasing**, respectively
* It has been quite hard to describe these UK trends, which I think is a **positive** thing. The comparison with other countries definitely helps with perspective


\


If you've enjoyed my analysis, **check out my recent XGBoost kernel [here](https://www.kaggle.com/lmorgan95/fraud-detection-smote-xgboost-business-impact)** and give me a follow to see future work!